---
title: "Introduction to R for Energy Data"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE)
```


**Hello class!**

To start our adventure in *R*, let's start by loading the required software software:

- [Download R](https://cloud.r-project.org/)
- [Download RStudio](http://www.rstudio.com/download)

Once you've done that, you want to go to RStudio and select:

- File > New File > RScript

You should now have a blank source pane in which to start writing!

Start by installing and loading packages. While R has base abilities, adding packages made by the myriad of users is what makes R so versatile. Let's add the `tidyverse` to start, it's a collection of important packages made by the people at RStudio. We should also add `readxl` because we are going to read in an excel file.

```{r, cache=TRUE}
#install.packages("tidyverse")  #this installs the package (remove the hashtag to uncomment and run this line the first time you run it)
#install.packages("readxl")  #this installs the package (remove the hashtag to uncomment and run this line the first time you run it)
library(tidyverse)  #this loads the installed package
library(readxl)  #this loads the installed package
```

Next, lets load in some simple data just to play around.

We will start by downloading a file from the AESO's Data Requests page (a treasure trove of great data sets!). Once that is downloaded, we will read it into R as a dataframe object called `data`:

```{r, cache=TRUE}
download.file("https://aeso.ca/assets/Uploads/data-requests/System-Load-Annual-2009-to-2022.xlsx",
              destfile="annual_load.xlsx",
              mode="wb")
data <- read_xlsx("annual_load.xlsx")
```

Let's take a peak at this data object:
```{r, cache=T}
head(data)
```
Ok! We have data! But there's a weird third column and i don't like the variable names. Let's delete the 3rd column and rename the first two. To do that we will create a new dataframe called `df` and use the `tidyverse` to make some changes:
```{r, cache=T}
df <- data %>%   #assign data to a new object called df and then (%>%)...
  select(-3)  %>% #select all columns that are NOT column 3 and then (%>%)...
  rename(year=Year,load=`System Load (GWh)`)  #rename the columns; note the backticks, that's how you deal with variable names that have spaces.
head(df)  #let's peek at the first few rows of this new object
```
Much nicer!

Now let's explore the data using a quick chart. Let's plot the annual system load by year.
```{r}
ggplot(df,aes(x=year,y=load))+  #we are creating a plot using the dataframe `df` and the aesthetics will involve year on the x-axis and load on the y-axis
  geom_line()  #plot a line using the aesthetics assigned above (note we could have assigned them here as well)
```
Not bad! Our first chart!
```{r}
ggplot(df,aes(year,load,label=load))+
  geom_col(fill="firebrick",alpha=.8)+
  geom_text(vjust=1.1,size=3,color="white")+
  scale_x_continuous(breaks=seq(2009,2022,1))+
  scale_y_continuous(limits=c(0,75000),expand=c(0,0))+
  labs(x="",y="",
       title="Alberta's Annual System Load",
       subtitle="Annual GWh",
       caption="Source: AESO.ca data; Chart by Blake")+
  theme_light(14)+
  theme(panel.grid.minor=element_blank(),
        panel.grid.major.x=element_blank(),
        axis.text.x=element_text(angle=90))
```







